---
title: "EF tasks markdown 2020 - EFA"
author: "Christoph Voelter"
date: "Octovber , 2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(tidyverse)
library(psych)
library(nFactors)
library(summarytools)
library(corrr)
library(reshape2)
library(ggthemes)
library(ggpubr)

```

### Load data

```{r}
EF_task_data<-read.csv("EFTB_CFA_data_chimpanzees.csv")
```

```{r}
efa.data<-EF_task_data%>%
  select(-X, -Subject,  -ED, -SD, -SR,-ID, )%>%
  drop_na%>%
  rename("WM Updating"="WMUpdating",
         "WM Boxes"="WMBoxes",
         "WM Grid"="WMGrid",
         "Inhibition Cylinders"="Cylinder",
         "Inhibition Grid"="InhibGrid",
         "Inhibition Boxes"="GlassCeiling",
         "Shifting Tray"="Tray",
         "Shifting Shelf"="Shelf",
         "Shifting Boxes"="CD_all")
efa.data<-efa.data%>%
  select(order(colnames(efa.data)))%>%
  select(-"WM Grid",-"WM Updating",  -"Inhibition Grid" )
```


Plotting DVs across sites

```{r}

efa.data.site<-read.csv(file="EF_task_individual_corr.csv")%>%
  select(-ED, -SD, -SR,-ID, -Group, -Sex, -Age)%>%
  drop_na%>%
  rename("WM_Updating"="Updating_P1P2.correct",
         "WM_Boxes"="boxes_correct_newrange",
         "WM_Grid"="Grid_Distance_newrange",
         "Inhibition_Grid"="Inh.grid.correct1",
         "Inhibition_Cylinders"="Inh.cylinders.correct1",
         "Inhibition_Boxes"="Inh.boxes.correct1",
         "Shifting_Tray"="Shifting.tray.correct.S2",
         "Shifting_Shelf"="proportion_switches",
         "Shifting_Boxes"="CD")



efa.data.plot<-efa.data.site%>%
  gather(task, performance, WM_Updating:Shifting_Boxes)%>%
  mutate(task=str_replace(task, "_", " "), site=fct_recode(site, "Sweetwaters"="kenya", "Ngamba"="uganda"))%>%
  group_by(task,site)%>%
  add_count(performance)%>%
  ungroup()

efa.data.ttests.site<-efa.data.site %>%
    summarise_each(funs(t.test(.[site == "kenya"], .[site == "uganda"], paired = FALSE, alternative = "two.sided")$p.value), vars = WM_Updating:Shifting_Boxes)

names(efa.data.ttests.site) <- colnames(efa.data.site[2:10])

site.plot<-ggplot(data=efa.data.plot, aes(x=site, y=performance))+
  geom_boxplot(alpha=0.5, outlier.colour = "white")+
  geom_point(color="grey", alpha=0.1, size=efa.data.plot$n)+
  facet_wrap(~task)+
  theme_few()+
  ylab("Standardised performance")+ xlab("Site")+
  stat_compare_means(paired = FALSE, method = "t.test", label.y = 0.95 )
  
site.plot

ggsave(site.plot, filename = "graphics/site_plot.png", height=10, width = 10, scale=0.65)

  
```



### Correlation matrix based on all participants and all tasks
```{r echo=FALSE, warning=FALSE}

EF_task_data2 <- read.csv("EFTB_CFA_data_chimpanzees.csv")%>%
  select(-X, -Subject)%>%
  rename("WM Updating"="WMUpdating",
         "WM Boxes"="WMBoxes",
         "WM Grid"="WMGrid",
         "Inhibition Cylinders"="Cylinder",
         "Inhibition Grid"="InhibGrid",
         "Inhibition Boxes"="GlassCeiling",
         "Shifting Tray"="Tray",
         "Shifting Shelf"="Shelf",
         "Shifting Boxes (CD)"="CD_all",
         "Shifting Boxes (SD)"="SD",
         "Shifting Boxes (SR)"="SR",
         "Shifting Boxes (ID)"="ID",
         "Shifting Boxes (ED)"="ED")%>%
  select(order(desc(colnames(.))))


  
library(ggcorrplot)
corr <- cor(EF_task_data2, use = "pairwise.complete.obs")
p.mat <- cor_pmat(EF_task_data2, use = "pairwise.complete.obs")
cormat.all.tasks<- ggcorrplot(corr, hc.order = FALSE,
  sig.level = 0.05,
  type = "lower",
  p.mat = p.mat,
  insig = c("pch"), pch = 4, pch.col = "darkgrey",
  pch.cex = 5,
  lab = TRUE,
  digits = 2
) +
  scale_fill_gradient2(
    low = "dodgerblue", high = "darkorange",
    midpoint = 0, limit = c(-1, 1)
  ) +
  guides(fill = F) +
  theme_few(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    plot.margin = unit(c(9, 1, 0, 0), "mm")
  )
cormat.all.tasks
ggsave(cormat.all.tasks, filename="graphics/cormat_all_tasks.png", width=10, height=10, scale=0.65)


```


# EFA

## Parallel analysis

```{r}

# Determine Number of Factors to Extract

ev <- eigen(cor(efa.data)) # get eigenvalues
ap <- parallel(subject=nrow(efa.data),var=ncol(efa.data),
  rep=1000,cent=.5) #parallel analysis using package nFactors
#This function gives the distribution of the eigenvalues of correlation or a covariance matrices of random uncorrelated standardized normal variables. The mean and a selected quantile of this distribution are returned.

nS <- nScree(x=ev$values, aparallel=ap$eigen$mevpea)#quantile of the eigenvalues distribution is used
plotnScree(nS) 

#parallel analysis using package psych
fa.parallel(efa.data, cor="cor",  fm="ml", fa="pc", n.iter = 1000, SMC=TRUE)
#For comparability with other algorithms (e.g, the paran function in the paran package), setting smc=TRUE will use Squared Multiple Correlation as estimates of communalities. 
```
## Scree plot for publication
```{r}
#credit: https://rdrr.io/github/Rkabacoff/factorAnalysis/src/R/screePlot.R
x <- fa.parallel(efa.data, cor = "cor", fm = "ml", fa = "pc", n.iter = 100, SMC = TRUE)

x$fa.simr
n.iter <- 100
actual <- x$pc.values
sim <- x$pc.sim
nvar <- ncol(efa.data)
nfact <- x$ncomp

actual <- data.frame(n = 1:nvar, eigenvalue <- actual, type = "actual")
sim <- data.frame(n = 1:nvar, eigenvalue <- sim, type = "simulated")
names(actual) <- names(sim) <- c("n", "eigenvalue", "type")
plotdata <- as.data.frame(rbind(actual, sim))

apatheme <- theme_bw() +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    # text=element_text(family='Arial'),
    legend.title = element_blank(),
    legend.position = c(.7, .8),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black")
  )

p <- ggplot(plotdata, aes(x = n, y = eigenvalue, shape = type, color = type)) +

  # Add lines connecting data points
  geom_line() +

  # Add the data points.
  geom_point(size = 4) +

  # Label axes and titles
  labs(
    y = "Eigenvalue",
    x = "Number of factors" # ,
    # title = "Scree Plot",
    # subtitle = paste0("with parallel analysis (", n.iter, " iterations)"),
    # caption = "Maximum Likelihood Factoring"
  ) +

  # Ensure x-axis ranges from 1 to max # of factors,
  # increasing by one with each 'tick' mark.
  scale_x_continuous(breaks = min(plotdata$n):max(plotdata$n)) +

  # Ensure y-axis  that it ranges from below the lowest eigenvalue
  #  to above the highest eigenvalue an increments by .25.
  scale_y_continuous(breaks = round(seq(
    from = floor(min(plotdata$eigenvalue)),
    to = ceiling(max(plotdata$eigenvalue)),
    by = .5
  ), 2)) +

  # Manually specify the different shapes to use for actual and simulated data,
  # in this case, white and black circles.
  scale_shape_manual(values = c(16, 1)) +

  # Add horizontal line indicating Kaiser Harris criteria
  geom_hline(yintercept = 1, linetype = "dashed", color = "lightgray") +

  # Add vertical line indicating parallel analysis suggested max # of factors to retain
  geom_vline(xintercept = nfact, linetype = "dashed") +
  apatheme

# return plot
ggsave(p, filename = "graphics/scree_plot_chimp.png", width = 6, height = 6, scale = 0.6)

```

### Bartlett's test of sphericity
```{r}
cortest.bartlett(efa.data)

```

### Kaiser-Meyer-Olkin factor adequacy
```{r}
#cylinder task excluded due to low Measure of sampling adequacy (<0.3)
KMO(efa.data)

```

### 2-factor model

```{r}
fit <- factanal(efa.data, 2, rotation="varimax", cutoff=0, sort=FALSE)
print(fit, digits=2, cutoff=0, sort=TRUE)
xx2<-fit$loadings[,]

summary(xx2)

fa.diagram(fit$loadings, digits=2,cut=.2, simple=FALSE,  adj = 3)

```
### 1-factor model

```{r}
fit1 <- factanal(efa.data, 1, rotation="varimax", cutoff=0, sort=FALSE)
print(fit1, digits=2, cutoff=0, sort=FALSE)
xx1<-fit1$loadings[,]

summary(xx1)

fa.diagram(fit1$loadings, digits=2,cut=.1, simple=FALSE,  adj = 3)

```

### Correlation matrix plot (based on all participants that completed all tasks)
```{r}
  get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
  }
efa.data.rev<-efa.data%>%
  select(order(desc(colnames(efa.data))))

cormat <- round(cor(efa.data.rev, method="pearson", use = "pairwise.complete.obs"),2)
upper_tri <- get_upper_tri(cormat)
melted_cormat=melt(cormat, na.rm = TRUE)

cor.plot<-ggplot(data = melted_cormat, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile()+
  geom_text(aes(Var2, Var1, label = value), color = "black", size = 2) +
 scale_fill_gradient2(low = "dodgerblue", high = "darkorange", 
   midpoint = 0, limit = c(-1,1)) +
  theme_few(base_size=10)+ 
 #theme(axis.text.x = element_text(angle = 45, vjust = 1, 
  #  size = 8, hjust = 1), legend.position = "none")+
  #labs(x="",y="")+
 #coord_fixed()
    theme(axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1), 
        axis.title.x=element_blank(), 
        axis.title.y=element_blank(), 
        plot.margin = unit(c(8, 1, 0, 0), "mm")) +
  guides(fill=F)

cor.plot
```
```{r}
library(ggcorrplot)

pvals<-cor_pmat(efa.data.rev)
ggcorrplot(cor(efa.data.rev, use = "pairwise.complete.obs"),
  type = "full",
  p.mat = pvals,
  insig = c("pch"), pch = 4, pch.col = "grey",
  pch.cex = 5,
  lab = TRUE,
  digits = 2
) +
  scale_fill_gradient2(
    low = "dodgerblue", high = "darkorange",
    midpoint = 0, limit = c(-1, 1)
  ) +
  guides(fill = F) +
  theme_few(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    plot.margin = unit(c(8.5, 1, 0, 0), "mm")
  )
```
### Plot: 2-factor model loadings
```{r}
loadings <- melt(xx2)

loadings<-loadings%>%
  rename(Measure="Var1", Factor="Var2")%>%
  arrange(Factor, Measure)

#2-factor solution
loading.plot_2fac<-ggplot(loadings, aes(as.factor(desc(Measure)), abs(value), fill=value)) + 
  facet_wrap(~ Factor, nrow=1) + #place the factors in separate facets
  geom_bar(stat="identity") + #make the bars
  coord_flip() + #flip the axes so the test names can be horizontal  
  #define the fill color gradient: blue=positive, red=negative
  scale_fill_gradient2(name = "Loading", 
                       low = "dodgerblue", mid = "white", high = "darkorange", 
                       midpoint=0, guide=F) +
  ylab("Loading Strength") + #improve y-axis label
  theme_few(base_size=10)+ #use a black-and0white theme with set font size
    theme(panel.spacing = unit(0.8, "lines"),
          axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        plot.margin = unit(c(2.3, 3, 12.3, 1), "mm")) 
```

### Plot: 1-factor model loadings

```{r}
loadings1 <- melt(xx1)


loadings1<-loadings1%>%
  add_rownames()%>%
  rename(Measure="rowname")%>%
  mutate(Factor="Factor 1")%>%
  mutate_if(is.character,as.factor)%>%
  arrange(Measure)

#2-factor solution
loading.plot_1fac<-ggplot(loadings1, aes(as.factor(desc(Measure)), abs(value), fill=value)) +   geom_bar(stat="identity") + #make the bars
    facet_wrap(~ Factor, nrow=1) + #place the factors in separate facets
  coord_flip() + #flip the axes so the test names can be horizontal  
  #define the fill color gradient: blue=positive, red=negative
  scale_fill_gradient2(name = "Loading", 
                       low = "dodgerblue", mid = "white", high = "darkorange", 
                       midpoint=0, guide=F) +
  ylim(0,1)+
  ylab("Loading Strength") + #improve y-axis label
  theme_few(base_size=10)+ #use a black-and0white theme with set font size
    theme(axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        plot.margin = unit(c(2.3, 3, 12.3, 1), "mm")) 

loading.plot_1fac
```


```{r}
library(cowplot)
efa_plot<-plot_grid(cor.plot, loading.plot_1fac, loading.plot_2fac, labels = c("A", "B", "C"), nrow=1, rel_widths = c(1.9,0.8, 1.6))
efa_plot
ggsave(efa_plot, filename = "graphics/efa_chimp_plot_1_and_2factors.png", width = 13, height=6, scale=.55)
```

